---
title: "final"
format: html
editor: visual
---

# Github link

[Github Repository](https://github.com/tanveersiingh/ENVS-193DS_spring-2025_final.git)

# Set up

```{r}
# load packages
library(tidyverse) # general use
library(here) # file organization
library(flextable) # creating tables
library(janitor) # cleaning data frames
library(officer) # table formatting
library(dplyr)
library(lubridate)
# read in data
sst <- read_csv("./data/SST_update2023.csv")
```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a Pearson correlation test.In part 2, they used a one way ANOVA test.

## b. More information needed

The one-way ANOVA shows that at least one source differs in average nitrogen load, but not which one(s). I recommend adding a Tukey's HSD post hoc test to determine which specific source pairs differ significantly. Additionally, although p = 0.02 indicates statistical significance, it does not show the magnitude of those differences. To address this, I suggest calculating Cohen’s d to quantify the effect size between each pair of sources, indicating whether differences are small, medium, or large.

## c. Suggestions for rewriting

Biological narrative: Average nitrogen load (kg year$^{-1}$) differs among sources, with urban land contributing notably higher nitrogen levels compared to grasslands, suggesting source-specific impacts on nitrogen pollution. Statistical summary: We rejected the null hypothesis that there is no difference in average nitrogen load (kg year$^{-1}$) among sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands). We found a small/medium/large effect size ($\eta^2$ = effect size) using a one-way ANOVA (F(df$_1$, df$_2$) = F-statistic, p = 0.02, $\alpha$ = significance level). On average, Source A has a higher/lower nitrogen load than Source B (difference between groups, 95% CI: \[lower, upper\] kg year$^{-1}$).

# Problem 2. Data visualization

```{r}
sst_clean <- sst |> # start with df
  mutate( # use mutate the create year and month from date
    year = year(date), # create year column
    month = month(date, label = TRUE, abbr = TRUE) # create month column with the abbreviations 
  ) |>
  filter(year >= 2018 & year <= 2023) |> # selecting only years 2018 - 2023
  mutate(year = as.factor(year)) |> # change year column to factor
  group_by(year, month) |> # grouping year and month for summary calculations
  summarize(mean_monthly_sst = round(mean(temp, na.rm = TRUE), 1)) # summarize and round mean_monthly_sst

print(slice_sample(sst_clean, n = 5)) # slice and print 5 random rows
print(str(sst_clean)) # print the structure of sst_clean
```

